---
title: "Easy use of Google Earth Engine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Easy use of Google Earth Engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(envfetch)
```

## Need

In cases where downloading of local files for extracting data are not possible 
or practical, cloud solutions can provide a performant alternative. The largest
and most commonly used cloud-based geospatial analysis platform is Google
Earth Engine.

Currently, there is only one R package with an implemented extraction function
for Google earth engine. However, with large datasets or extraction tasks,
standard use of this function can causes crashes or overuse of Google Earth
Engine's memory limits, failing the extraction task. The `envfetch` package 
implements the `extract_gee` function to address this problem.

## Overuse of GEE memory limits with `rgee`

## Completion of task by respecting GEE memory limits with the `extract_gee` package
